Predicting Music Emotion: Insights from Canadian Weather and Spotify Audio Features

### 2024-12-04 paper/paper.qmd
final changes. done.

---
title: "My title"
subtitle: "My subtitle if needed"
author: Cristina Su Lam
thanks: "Code and data are available at: https://github.com/cristinaasu/Spotify_Analysis"
date: today
date-format: long
abstract: "This analysis explores the relationship between various musical and contextual features with the emotional tone (Valence) of songs on Spotify. Using a dataset of songs with features such as tempo, danceability, and acousticness, alongside contextual data like mean temperature, we aim to predict Valence—a measure of positivity in the music. The study applies multiple linear regression models, evaluates performance metrics, and identifies influential predictors. Results show that predictors like danceability and energy strongly relate to Valence, while contextual variables like temperature also provide moderate explanatory power. These findings highlight the potential for combining musical and environmental data to analyze listener sentiment and mood-related trends in music[need to be more detail about findings]."
toc: true
fig_caption: yes
number-sections: true
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(hexbin)
library(arrow)
library(kableExtra)
library(knitr)
```

```{r}
# Load dataset 
analysis_data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
mlm_model <- readRDS("../models/mlm_model.rds")
```

# Introduction

Valence, a measure of emotional tone in music, is pivotal for understanding how listeners perceive and engage with songs. Ranging from 0 (negative or sad) to 1 (positive or happy), Valence provides insights into the emotive qualities of a track. This study aims to model and predict Valence using a blend of musical features such as tempo, danceability, and acousticness, alongside contextual variables like mean temperature.

Music serves as a universal language, evoking emotions and creating connections. By analyzing how intrinsic musical properties interact with external factors, such as environmental conditions, we can uncover patterns that shape listener sentiment. This approach not only deepens our understanding of the factors influencing emotional tone but also offers practical applications for playlist curation and music recommendation systems.

The estimand of interest is Valence, as it encapsulates the core emotional response elicited by music. Through multiple linear regression (MLR) models, we identify the most significant predictors of Valence and assess their explanatory power. This approach highlights the interplay between quantitative attributes of music and listener experiences, aiming to improve the accuracy of mood prediction models.

In this analysis, we seek to bridge the gap between musical analytics and contextual data, contributing to the broader field of musicology and data-driven sentiment analysis. By refining predictive methods, we aim to better understand the emotional landscape of modern music and its relationship with the listener's environment.

\[Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....\]

@@

@@

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

The dataset contains key variables spanning musical features, environmental data, and outcomes. Below is a summary of numerical variables:

```{r}
#| label: tbl-summary
#| tbl-cap: Summary Statistics
#| echo: false
#| warning: false
#| fig-align: center
summary_table <- analysis_data %>%
  reframe(
    Variable = c("Valence", "Mean Temp", "Danceability", "Energy", "Acousticness", "Tempo"),
    Mean = c(
      mean(Valence, na.rm = TRUE),
      mean(`Mean Temp`, na.rm = TRUE),
      mean(Danceability, na.rm = TRUE),
      mean(Energy, na.rm = TRUE),
      mean(Acousticness, na.rm = TRUE),
      mean(Tempo, na.rm = TRUE)
    ),
    SD = c(
      sd(Valence, na.rm = TRUE),
      sd(`Mean Temp`, na.rm = TRUE),
      sd(Danceability, na.rm = TRUE),
      sd(Energy, na.rm = TRUE),
      sd(Acousticness, na.rm = TRUE),
      sd(Tempo, na.rm = TRUE)
    ),
    Min = c(
      min(Valence, na.rm = TRUE),
      min(`Mean Temp`, na.rm = TRUE),
      min(Danceability, na.rm = TRUE),
      min(Energy, na.rm = TRUE),
      min(Acousticness, na.rm = TRUE),
      min(Tempo, na.rm = TRUE)
    ),
    Max = c(
      max(Valence, na.rm = TRUE),
      max(`Mean Temp`, na.rm = TRUE),
      max(Danceability, na.rm = TRUE),
      max(Energy, na.rm = TRUE),
      max(Acousticness, na.rm = TRUE),
      max(Tempo, na.rm = TRUE)
    )
  ) %>%
  mutate(
    Mean = round(Mean, 3),
    SD = round(SD, 3),
    Min = round(Min, 3),
    Max = round(Max, 3)
  )
kable(summary_table, format = "pipe")
```

These metrics reveal diverse song features and moderate variability in environmental temperature, offering rich explanatory potential for Valence.

The data used in this analysis were collected from Spotify's public API, which provides detailed metadata for songs. Each track is associated with multiple audio features and contextual information. The data was preprocessed to ensure consistency, with missing or erroneous values removed prior to analysis.

Valence (Outcome): Extracted directly from Spotify's API, Valence is measured on a scale from 0 to 1, indicating the positivity or happiness conveyed by a song. This variable serves as the primary outcome of interest.

Musical Features:

Danceability: Quantifies how suitable a track is for dancing, based on a combination of tempo, rhythm stability, and beat strength. Values range from 0 to 1.

Energy: Measures the intensity and activity level of a track, combining attributes like loudness and dynamic range. Higher values indicate more energetic tracks.

Acousticness: Represents the likelihood of a track being acoustic, with values closer to 1 reflecting higher acoustic qualities.

Tempo: Captured in beats per minute (BPM), tempo describes the overall speed of a track. This feature was extracted directly from Spotify's analysis tools.

Contextual Feature:

Mean Temp: Weather data was integrated using external sources and mapped to listening periods. This feature reflects the average daily temperature (°C) corresponding to the time a track was popular or listened to.

The integration of audio and contextual data allowed for a comprehensive analysis of factors influencing Valence, bridging the gap between intrinsic musical properties and external environmental conditions.

\[Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.\]

## Outcome variable

The distribution of Valence shows a relatively uniform spread, with some concentration near the mid-range. This reflects the diversity of emotional tones in the dataset. The histogram of Valence illustrates this distribution, where most tracks lie between moderate to high levels of positivity.

```{r}
#| label: fig-valence
#| fig-cap: Distribution of Valence
#| echo: false
#| warning: false
#| fig-align: center

ggplot(analysis_data, aes(x = Valence)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(x = "Valence", y = "Count") +
  theme_minimal()
```

## Predictor variables

### Mean Temperature

The average temperature over time demonstrates a seasonal decline, ranging from approximately 25°C in the summer to below 10°C in late autumn. This variable adds a contextual layer, capturing external influences that may correlate with song Valence and listening behavior.

```{r}
#| label: fig-temp
#| fig-cap: Average temperature Over Time
#| echo: false
#| warning: false
#| fig-align: center
ggplot(analysis_data, aes(x = Date, y = `Mean Temp`)) +
  geom_line(color = "red") +
  labs(title = "Average Temperature Over Time", x = "Date", y = "Temperature (°C)") +
  theme_minimal()
```

### Tempo

The distribution of Tempo reveals a central peak around 120 BPM, with a wider spread for slower and faster songs. This suggests a balanced mix of tempos, with a slight preference for moderate-speed tracks. Tempo's role in influencing Valence is analyzed alongside other musical features.

```{r}
#| label: fig-tempo
#| fig-cap: Distribution of Tempo
#| echo: false
#| warning: false
#| fig-align: center
ggplot(analysis_data, aes(x = Tempo)) +
  geom_histogram(binwidth = 10, fill = "darkgreen", color = "black") +
  labs(title = "Distribution of Tempo", x = "Tempo (BPM)", y = "Frequency") +
  theme_minimal()
```

### Danceability, Energy, and Acousticness

The comparative histograms of Danceability, Energy, and Acousticness highlight their distinct distributions:

Danceability: Peaks near 0.6 to 0.8, suggesting that most songs are moderately to highly danceable.

Energy: Similarly peaks in the mid-range, reflecting a tendency towards active, engaging tracks.

Acousticness: Displays a skewed distribution, with most tracks having low acoustic properties, indicating a focus on electronic or produced music styles.

Each of these variables provides unique explanatory power, contributing to the overall understanding of what drives the emotional tone in music.

```{r}
#| label: fig-diff
#| fig-cap: Distribution of Tempo
#| echo: false
#| warning: false
#| fig-align: center
analysis_data %>%
  select(Danceability, Energy, Acousticness) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value, fill = Variable)) +
  geom_histogram(binwidth = 0.05, alpha = 0.6, position = "identity", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  labs(
    title = "Distribution of Danceability, Energy, and Acousticness",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

```{=text}
\begin{align}
y_i &= \beta_0 + \beta_1 (\text{Mean Temp}_i) + \beta_2 (\text{Artist}_i) + \beta_3 (\text{Danceability}_i) \\
&\quad + \beta_4 (\text{Energy}_i) + \beta_5 (\text{Acousticness}_i) + \beta_6 (\text{Tempo}_i) + \epsilon_i \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}
```
- $y_i$: 
- $\beta_0$:
- $\beta_1$: 
- ... [add beta's for the other predictors]
- ...
- $u_{\text{pollster}}$:
- $u_{\text{state}}$:

### Model justification

```{r}
#| label: tbl-comp
#| fig-cap: Comparison of Model Performance
#| echo: false
#| warning: false
#| fig-align: center
comparison_data <- data.frame(
  Model = c("MLR 1", "MLR 2", "MLR 3"),
  `# Variables` = c(6, 5, 4),
  RMSE = c(0.0846, 0.0866, 0.2031),
  `R-Squared` = c(0.8676, 0.8600, -1.4266),
  check.names = FALSE
)

kable(comparison_data, format = "pipe")
```

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "Predicting Music Emotion: Insights from Canadian Weather and Spotify Audio Features"
subtitle: "Danceability Dominates, while Temperature Plays a Minor Role in Shaping Positivity in the Top 50 Charts"
author: Cristina Su Lam
thanks: "Code and data are available at: https://github.com/cristinaasu/Spotify_Analysis"
date: today
date-format: long
abstract: "This study explores how musical features and temperature shape emotional perceptions of music, measured by valence—a scale capturing a track's positivity. Using Spotify’s top 50 Canadian charts and population-weighted mean temperatures from six major cities, we applied a multiple linear regression model. Danceability emerged as the strongest predictor of valence, with an effect size of 1.260, while scaled temperature had a smaller but significant effect of 0.007. These findings highlight the interplay between intrinsic audio features and environmental contexts in influencing listener sentiment, offering insights into music's emotional dynamics and practical applications for personalized playlists, mood-based marketing, and further research on cultural and environmental influences."
toc: true
fig_caption: yes
number-sections: true
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(hexbin)
library(arrow)
library(kableExtra)
library(knitr)
library(modelsummary)
library(broom)
library(car)
```

```{r}
#| include: false
#| warning: false
#| message: false
# Load dataset 
analysis_data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
mlm_model <- readRDS("../models/mlm_model.rds")
```

# Introduction {#sec-intro}

Music is a deeply embedded and transformative part of people’s lives, offering experiences that range from solitary headphone listening on a chilly day to attending global live events. It fosters empathy, social bonding, and cultural understanding, connecting individuals across cultural and linguistic differences [@CLARKE201561]. A UK study by @anglada2023music revealed that musical preferences are influenced by weather; warmer temperatures encourage songs with high intensity and positive emotions, while increased rainfall sees a decline in their popularity. This underscores how weather significantly shapes the musical landscape.

The digital age has provided unprecedented access to music data, with platforms like Spotify offering rich datasets on song characteristics [@spotifyr]. Inspired by the previous research, this study aims to bridge the gap by examining the interplay of musical and environmental factors on listener emotions within the Canadian context. Using valence—a metric measuring emotions from 0 (negative or sad) to 1 (positive or happy)—the study evaluates how these elements collectively shape listener experiences.

Valence, the estimand of interest in this study, represents the true effect this research seeks to understand. While it itself is unobservable, this study estimates it through a structured analytical approach, ensuring that the methodology remains sharply focused on the central research question. As an abstract quantity, it captures the theoretical relationship independent of sample-specific variations, allowing the study to delineate true effects from noise and systematic biases inherent in observational data.

Using a multiple linear regression framework, this study analyzed data from Spotify’s Top 50 Canadian charts, combined with temperature data from six major cities, spanning June to November 2024. Danceability emerged as the strongest predictor of valence, underscoring its dominant influence on listener emotions. Artist-specific effects, such as the positive contributions of Brenda Lee's tracks and the negative impact of Future's music, were also notable. Interestingly, temperature—despite being the primary variable of interest—had a relatively minor influence, suggesting that weather exerts a limited direct effect on valence. These findings highlight the need to integrate contextual and musical factors to design personalized recommendations and mood-based marketing strategies that resonate with diverse listener preferences.

The remainder of this paper is structured as follows. @sec-data describes the dataset and measurement process, while @sec-model outlines the methodology and model validation. @sec-results presents the results, highlighting the key predictors of valence and their implications. Finally, @sec-discussion discusses the findings, limitations, and potential avenues for future research.

@@

## Overview {#sec-overview}

To analyze the relationship between song characteristics, temperature, and valence--data from Spotify and weather records from Canada's most populous cities were integrated. To be more specific, the dataset included daily Top 50 songs in Canada from Spotify Charts [@spotify_charts], covering the period from June 1, 2024, to November 16, 2024. Using the Spotify API [@spotifyr], features such as Valence, Danceability, Acousticness, and Tempo were retrieved. Concurrently, weather data was obtained from @historical_weather_gc for six major cities: Toronto, Montreal, Edmonton, Calgary, Ottawa, and Vancouver.

Data cleaning was conducted in R [@citeR] using the tidyverse [@thereferencecanbewhatever] and arrow [@arrow] packages. The cities mentioned were selected due to their large populations, and a new numerical variable, Mean Temp, was created by weighting each city's mean daily temperature by its population density and dividing by the total population density. This weighting was applied to account for the varying influence of each city based on its population size, ensuring that the derived temperature better reflects the overall climate experienced by the majority of the population. The resulting dataset consists of 8,450 observations and includes key variables such as Valence, Danceability, Acousticness, Tempo, Temperature, and Artist. The dataset was thoroughly checked for missing or incomplete values, confirming that all entries were complete and suitable for analysis. A sample of the cleaned dataset is displayed in [@tbl-spotifysongs].

For visualization and analysis, the knitr [@knitr], ggplot2 [@ggplot2], and kableExtra [@kableExtra] packages were utilized to generate professional tables and visualizations, effectively enhancing the presentation of results.

```{r}
#| label: tbl-spotifysongs
#| tbl-cap: Sample of Song's Characteristics
#| echo: false
#| warning: false
#| fig-align: center
analysis_data |> 
  select(Artist, Song, Valence, `Mean Temp`, Danceability, Acousticness, Tempo, Date) |> 
  mutate(
    `Mean Temp` = round(`Mean Temp`, 2),
    Acousticness = round(Acousticness, 3),
    Date = format(as.Date(Date), "%b %d") 
  ) |> 
  head(5) |> 
  kable(
    col.names = c("Artist", "Song", "Valence", "Mean Temp", "Dance", "Acoustic", "Tempo", "Date"),
  ) |> 
  kable_styling(
    full_width = FALSE, 
    position = "center", 
    font_size = 10
  ) |> 
  column_spec(1, width = "2.5cm") |>  
  column_spec(2, width = "2.7cm") |> 
  column_spec(3:8, width = "1.25cm") 
```

[@tbl-summary] presents the summary statistics of the numerical variables, revealing key insights into the dataset's characteristics and patterns. Valence, ranging from 0 to 1, spans 0.036 to 0.981 with a mean of 0.532, reflecting an overall moderate mood among tracks. Mean Temperature, calculated as a population-weighted average across major Canadian cities, has a mean of 16.9°C, with values spanning from 3.5°C to 24.8°C, showcasing the climatic variation during the study period.

Danceability and Acousticness, both measured on a 0 to 1 scale, highlight contrasting patterns. Danceability values range from 0.234 to 0.943, with a mean of 0.632, indicating that most songs are moderately to highly danceable. Acousticness, ranging from 0.000 to 0.968, has a lower mean of 0.228, suggesting that non-acoustic tracks---those with more electronic or synthetic instrumentation---dominate the dataset.

Finally, Tempo, measured in beats per minute (BPM), has a mean of 126 BPM, with a wide range from 48.7 BPM for slower tracks to 203.8 BPM for faster ones. This diversity in tempo underscores the variety of musical styles, forming a robust basis for exploring their relationship with valence.

```{r}
#| label: tbl-summary
#| tbl-cap: Summary Statistics
#| echo: false
#| warning: false
#| fig-align: center
summary_table <- analysis_data %>%
  reframe(
    Variable = c("Valence", "Mean Temp", "Danceability", "Acousticness", "Tempo"),
    Mean = c(
      mean(Valence, na.rm = TRUE),
      mean(`Mean Temp`, na.rm = TRUE),
      mean(Danceability, na.rm = TRUE),
      mean(Acousticness, na.rm = TRUE),
      mean(Tempo, na.rm = TRUE)
    ),
    SD = c(
      sd(Valence, na.rm = TRUE),
      sd(`Mean Temp`, na.rm = TRUE),
      sd(Danceability, na.rm = TRUE),
      sd(Acousticness, na.rm = TRUE),
      sd(Tempo, na.rm = TRUE)
    ),
    Min = c(
      min(Valence, na.rm = TRUE),
      min(`Mean Temp`, na.rm = TRUE),
      min(Danceability, na.rm = TRUE),
      min(Acousticness, na.rm = TRUE),
      min(Tempo, na.rm = TRUE)
    ),
    Max = c(
      max(Valence, na.rm = TRUE),
      max(`Mean Temp`, na.rm = TRUE),
      max(Danceability, na.rm = TRUE),
      max(Acousticness, na.rm = TRUE),
      max(Tempo, na.rm = TRUE)
    )
  ) %>%
  mutate(
    Mean = round(Mean, 3),
    SD = round(SD, 3),
    Min = round(Min, 3),
    Max = round(Max, 3)
  )
kable(summary_table, format = "pipe")
```

As noted in @sec-overview, the weather data originates from @historical_weather_gc, which collects meteorological observations from a comprehensive network of weather stations across the country. These stations measure key variables such as temperature, precipitation, and wind speed using standardized instruments and methodologies. Data is captured hourly or daily, depending on the station, and is validated through automated quality control processes and manual verification. This rigorous approach ensures reliable and precise measurements that meet international meteorological standards. In the context of this analysis, one weather station per city was selected based on relevance and proximity; additional details on this selection process are provided in @sec-appendix-a.

Similarly, the music data originates from Spotify Charts [@spotify_charts], which rank the most-streamed songs in a given region daily. These rankings are compiled by aggregating anonymized user streaming data across its platform. Streams are verified to ensure authenticity, eliminating artificial plays generated by bots or fraudulent activity. Additionally, Spotify employs machine learning models to analyze the audio features of each song, such as Valence, Danceability, Acousticness, and Tempo. These features are derived from the acoustic and digital signal processing of tracks, providing a quantitative representation of each song's characteristics.

It is important to note that in this dataset, daily temperature values are repeated across all fifty songs for a given day, as they represent an average measurement for the day rather than song-specific conditions. Similarly, artist entries recur within the dataset, reflecting the daily rankings of the Top 50 streamed songs, which often include repeated tracks for artists with sustained popularity or multiple charting songs.

The combination of these trusted data sources ensures that the dataset used in this study is grounded in accurate and validated measurements. Weather data reflects real-world environmental conditions, while Spotify's data captures both user engagement and precise audio analysis, enabling a comprehensive analysis of the relationship between temperature, song features, and valence.

## Analysis of Variables {#sec-analysis}

### Outcome variable

The Valence variable exhibits a fairly uniform distribution, as shown in [@fig-valence], with values ranging from 0.036 to 0.981, as detailed in [@tbl-summary]. This wide range captures a broad spectrum of emotional tones in the dataset.

```{r}
#| label: fig-valence
#| fig-cap: Uniform Distribution of Valence
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 5
#| fig.height: 3

ggplot(analysis_data, aes(x = Valence)) +
  geom_histogram(binwidth = 0.05, fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(x = "Valence", y = "Count") +
  theme_minimal()
```

[@tbl-highest] highlights the Top 5 songs with the highest Valence, led by "September" by Earth, Wind & Fire (0.981), alongside other highly positive tracks like "Apple" by Charli XCX (0.962) and "Super Freak" by Rick James (0.962). These songs represent the most uplifting and positive moods in the dataset.

```{r}
#| label: tbl-highest
#| tbl-cap: Top 5 songs with the Highest valence
#| echo: false
#| warning: false
#| tbl-align: center
# Show the top 5 songs with the highest Valence
top_valence_songs_unique <- analysis_data %>%
  distinct(Song, Artist, Valence) %>%
  arrange(desc(Valence)) %>%
  head(5)

kable(top_valence_songs_unique, col.names = c("Song", "Artist", "Valence"))
```

In contrast, [@tbl-lowest] features the Top 5 songs with the lowest Valence, including "BLUE" by Billie Eilish (0.0365) and "Meteor Man" by Lil Uzi Vert (0.0384), which reflect darker or more somber tones. Together, these extremes illustrate the dataset's diversity, capturing a broad emotional spectrum across tracks.

```{r}
#| label: tbl-lowest
#| tbl-cap: Top 5 Songs with the Lowest Valence
#| echo: false
#| warning: false
#| tbl-align: center
# Show the top 5 unique songs with the lowest Valence
lowest_valence_songs_unique <- analysis_data %>%
  distinct(Song, Artist, Valence) %>% 
  arrange(Valence) %>%
  head(5)

kable(lowest_valence_songs_unique, col.names = c("Song", "Artist", "Valence"))
```

### Predictor variables

-   **Mean Temperature**

[@fig-temp] shows the average temperature over time, demonstrating a seasonal decline from approximately 25°C in the summer to below 5°C in late autumn. This pattern aligns with expected seasonal transitions in Canada and is further supported by [@tbl-summary], where the Mean Temperature averages 16.9°C with a standard deviation of 5.298°C. This notable variability provides an opportunity to investigate potential correlations between environmental factors, such as weather, and shifts in musical preferences reflected in Valence.
```{r}
#| label: fig-temp
#| fig-cap: Average Temperature from June to November 2024
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 5
#| fig.height: 3

ggplot(analysis_data, aes(x = Date, y = `Mean Temp`)) +
  geom_line(color = "darkblue", alpha = 0.7) +
  labs(x = "Date", y = "Temperature (°C)") +
  theme_minimal()
```

\newpage
-   **Artist**

The variable Artist reveals intriguing patterns in the dataset. Displayed in [@fig-artist], the top 10 artists, led by Sabrina Carpenter and Zach Bryan, contributed the highest number of entries to the dataset, with their songs appearing repeatedly in the daily Top 50 rankings, each exceeding 600 occurrences. This underscores their strong presence on the charts during the study period. However, despite their prominence, their songs do not feature in [@tbl-highest] or [@tbl-lowest]. In contrast, artists like Chappell Roan and Billie Eilish reflect a broader emotional range, appearing in the highest and lowest Valence categories, respectively. Hozier, contributing the fewest entries among the Top 10, stands out for appearing in [@tbl-highest], showcasing the emotional positivity of some of his music despite a smaller overall presence.
```{r}
#| label: fig-artist
#| fig-cap: Sabrina Carpenter and Zach Bryan leads the Top 10 artists
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 5
#| fig.height: 3
# Get the top 10 artists by song count
top_artists <- analysis_data %>%
  count(Artist, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  mutate(Artist = gsub(" ", "\n", Artist))

ggplot(top_artists, aes(x = reorder(Artist, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
  labs(x = "Artist",
       y = "Number of Tracks") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0.5, size = 6))
```

-   **Tempo**

The distribution of tempo values, as depicted in [@fig-tempo], shows a roughly normal shape, with most tracks falling between 100 and 130 beats per minute (BPM) and peaking around 120 BPM. This aligns with findings that today's popular music often falls within the 100–140 BPM range [@masterclass_bpm], highlighting rhythmic patterns that resonate with mainstream listener preferences.

In addition, the distribution reveals outliers on both ends of the spectrum, with slower tempos below 60 BPM and faster tempos above 180 BPM representing niche genres or unique musical styles.

```{r}
#| label: fig-tempo
#| fig-cap: Normal Distribution of Tempo
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 5
#| fig.height: 3

ggplot(analysis_data, aes(x = Tempo)) +
  geom_histogram(binwidth = 10, fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(x = "Tempo (BPM)", y = "Frequency") +
  theme_minimal()
```
\newpage
-   **Danceability and Acousticness**

The [@fig-diff] below highlights distinct patterns in the distributions of Danceability and Acousticness.

Acousticness displays a strongly right-skewed distribution, with most values concentrated near 0, reaffirming the prevalence of non-acoustic tracks in the dataset as noted in [@sec-overview]. This pattern aligns with its low mean of 0.228 and relatively high variability, as shown in [@tbl-summary], capturing a mix of purely electronic tracks and a smaller presence of acoustic elements.

In contrast, Danceability demonstrates a more uniform distribution, peaking between 0.6 and 0.8. With a higher mean of 0.632 and a standard deviation of 0.134, this indicates that the majority of tracks are moderately to highly danceable. This distribution underscores the dataset's emphasis on popular music, where rhythmic appeal and danceability are often prioritized.

```{r}
#| label: fig-diff
#| fig-cap: Acousticness highlights tracks with lower acoustic qualities, while Danceability shows a balanced distribution across rhythmic and energetic levels.
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 6
#| fig.height: 4
#| fig.pos: "H" 
analysis_data %>%
  select(Danceability, Acousticness) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value, fill = Variable)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7, position = "identity", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  scale_fill_manual(values = c("Danceability" = "#4B0082", "Acousticness" = "#008080")) + # Custom colors
  labs(
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Exploring Relationships

-   **Valence vs. Temperature**

The heatmap in [@fig-heatmap] illustrates that higher temperatures (20-25°C) are associated with a greater frequency of tracks exhibiting moderately positive valence values, clustering around 0.5-0.75. In contrast, at lower temperatures (5-15°C), songs are distributed more evenly between positive and negative valence values. Notably, the highest frequency is observed at a valence of approximately 0.25 when temperatures range between 15-20°C. The absence of data in the lowest temperature range (0-5°C) reflects a limitation of the dataset rather than an actual trend.
\newpage
```{r}
#| label: fig-heatmap
#| fig-cap: Higher temperatures align with positive valence values, while lower temperatures show a wider spread in emotional tones.
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 5.5
#| fig.height: 3.5
#| fig.pos: "H" 
analysis_data %>%
  mutate(Temp_Range = cut(`Mean Temp`, breaks = seq(0, 25, by = 5))) %>%
  ggplot(aes(x = Temp_Range, y = Valence, fill = after_stat(count))) + 
  stat_bin2d(binwidth = c(1, 0.1)) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
  labs(x = "Temperature Range (°C)",
    y = "Valence (Happiness)"
  ) +
  theme_minimal() +
  annotate(
    "text", x = 1, y = 0.05, label = "No data in this region", color = "darkred", size = 2.5
  ) +
  theme (
    legend.key.size = unit(0.4, "cm"), 
    legend.text = element_text(size = 7), 
    legend.title = element_text(size = 8) 
  )
```

-   **Valence vs. Artist**

The chart in [@fig-avgvalence] highlights notable contrasts in the emotional tone of music produced by the Top 10 artists. Hozier, Tommy Richman, and Chappell Roan stand out with the highest average Valence scores, suggesting their tracks lean toward more positive and uplifting moods. This trend aligns with their reputations for creating emotionally resonant music. Conversely, Kendrick Lamar and Zach Bryan, who rank lowest in average Valence, demonstrate a preference for deeper or more introspective themes, reflecting their storytelling-driven styles.
\newpage
```{r}
#| label: fig-avgvalence
#| fig-cap: Hozier leads with the most positive emotional tone, while Kendrick Lamar features the lowest average valence among the top contributors.
#| echo: false
#| warning: false
#| fig-align: center
#| fig.width: 5.5
#| fig.height: 3.5
#| fig.pos: "H" 
# Average Valence by Artist (Top 10 Artists)
top_artists <- analysis_data %>%
  group_by(Artist) %>%
  summarize(avg_valence = mean(Valence), count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  mutate(Artist = gsub(" ", "\n", Artist))

ggplot(top_artists, aes(x = reorder(Artist, avg_valence), y = avg_valence, fill = avg_valence)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Artist",
    y = "Average Valence"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(hjust = 0.5, size = 6))
```

# Model {#sec-model}

@@

This section presents a multiple linear regression (MLR) model to predict song valence based on audio features and contextual factors. It evaluates how danceability, acousticness, scaled tempo, scaled mean temperature, and artist-specific effects influence valence, offering a clear framework for analyzing these relationships.

The analysis utilized the R packages stats [@citeR], MLmetrics [@MLmetrics], and car [@car]. The stats package enabled MLR modeling with the `lm` function, while MLmetrics calculated RMSE and R-squared for model evaluation. The car package assessed multicollinearity using Variance Inflation Factors (VIF).

@@

Mathematical Representation:

\begin{align}
y_i &= \beta_0 + \beta_1 (\text{Scaled Mean Temp})_i + \beta_2 (\text{Artist})_i + \beta_3 (\text{Danceability})_i \\
&\quad + \beta_4 (\text{Acousticness})_i + \beta_5 (\text{Scaled Tempo})_i + \epsilon_i
\end{align}

-   $y_i$: Valence of the song, measuring perceived positivity.
-   $\beta_0$: Baseline valence when predictors are at their reference levels.
-   $\beta_1$: Effect of scaled mean temperature on valence.
-   $\beta_2$: Artist-specific effects on valence.
-   $\beta_3$: Effect of danceability on valence.
-   $\beta_4$: Effect of acousticness on valence.
-   $\beta_5$: Effect of scaled tempo on valence.
-   $\epsilon_i$: Unexplained variability, assumed to follow a normal distribution with mean 0 and variance.

## Variable Selection {#sec-variable}

In constructing the model, careful consideration was given to selecting predictors that accurately capture the relationships influencing song valence:

-   *Scaled Mean Temperature*: Represents the influence of environmental factors on emotional responses to music. Scaling normalizes its range for better interpretability and comparability across predictors.

-   *Artist*: Captures the unique stylistic and emotional contributions of individual artists, as certain artists consistently produce music associated with higher or lower valence.

-   *Danceability*: Captures a song's rhythm and energy, with higher values often evoking happiness and positivity, thereby increasing valence.

-   *Acousticness*: Higher acousticness often linked to calmer, introspective tones that may lower valence, while lower acousticness corresponds to more electronic or dynamic tracks with higher positivity.

-   *Scaled Tempo*: Reflects a song's speed, influencing its energy and emotional intensity, with faster tempos often enhancing valence. Scaling ensures consistency with other predictors, like Scaled Mean Temperature.

## Model Validation

The model validation process involved out-of-sample testing, where the dataset was split into training and testing sets in a 70-30 ratio. Three variations of the multiple linear regression model were tested, differing in the number of predictors included. Model performance was evaluated using Root Mean Square Error (RMSE) and R-squared values. As shown in [@tbl-comp], the models demonstrate varying degrees of predictive accuracy and explanatory power.

```{r}
#| label: tbl-comp
#| tbl-cap: Comparison of Model Performance
#| echo: false
#| warning: false
#| tbl-align: center

comparison_data <- data.frame(
  Model = c("MLR 1", "MLR 2", "MLR 3"),
  Variables = c(
    "Scaled Mean Temp, Artist, Danceability, Acousticness, Scaled Tempo, Date",
    "Scaled Mean Temp, Artist, Danceability, Acousticness, Scaled Tempo",
    "Scaled Mean Temp, Danceability, Acousticness, Scaled Tempo"
  ),
  RMSE = c(0.0868, 0.0866, 0.2031),
  `R-Squared` = c(0.8597, 0.8601, -1.4267),
  check.names = FALSE
)

kable(
  comparison_data,
  format = "pipe",
  align = "l",
  col.names = c("Model", "Variables", "R-Squared", "RSME")
) |> 
  kable_styling(
    full_width = FALSE, 
    position = "center", 
    font_size = 10
  ) |> 
  column_spec(1, width = "3cm") |>  
  column_spec(2, width = "7cm") |> 
  column_spec(3:4, width = "2cm")

@@

After not taking into consideration Artist, MLR 3 includes fewer predictors, it performs significantly worse, as evidenced by its negative R-squared and much higher RMSE, indicating poor fit and predictive ability. On the other hand, MLR 2 demonstrates a slight performance advantage over MLR 1 in both RMSE and R-squared metrics.

## Model Justification

The choice of MLR 2 is justified by its balance of predictive performance and model simplicity. The `Date` variable was excluded because it is indirectly related to `Scaled Mean Temp`, which already captures the temporal context more meaningfully, making its inclusion redundant and unnecessarily complex. Furthermore, this model includes the key variables of interest outlined in @sec-variable, ensuring it addresses the research objectives effectively. Lastly, MLR 2 slightly outperforms MLR 1 in terms of RMSE and R-squared, further supporting its selection as the final model for predicting song valence.

## Model Diagnostics
